BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-04-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
cleaning up manuscript

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.  These factors ensure that the specimens remain the premier botanical data source.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observed that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Evidently, apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
It rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, it does not include data cleaning functionalities beyond verifying nations of collection. 
While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.

@@

The spatial module has five required functions and two optional functions.

*coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data. political_grabber determines many levels of administrative ownership, including land management and public land survey system sections.  
*physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
*political_grabber*  grabs a wide range of administrative information for a collection site. 
*site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
*directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages. 
*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569). 
*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.

@@

The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.

@@

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.

@@

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)`records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.

In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.

Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species). 
It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error. 
The final record was an egregious error where the order of the specific epithet and the genus name.

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate (@patten2024geographic).
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.

Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users. 
While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.

While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications to project branches.

Accessioning often relies on the use of the Office Suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats.

>>>>

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.  These factors ensure that the specimens remain the premier botanical data source into perpetuity.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Evidently apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities.  
While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github Pages (https://sagesteppe.github.io/BarnebyLives/) and around 100 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.

@@

The spatial module has five required functions and two optional functions.  
*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.  
*coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data. 
*political_grabber* determines many levels of administrative ownership, including land management and public land survey system sections.  
*physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
*site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
*directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages. 
*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).

Please note that the function *physical_grabber* is the one portion of the package where a decoupling may exist between the collection site, and the resolution of the spatial data. 
While we expect the mismatch to be negligible for all effective purposes relating to: elevation, major geology type, and in general aspect, estimates of slope at this resolution may be biased - generally to lower angles.
For these reasons collectors must always make notes on the truly local environment which taxa are found in, and consider that the notes from BL reflect the greater landscape which a microfeature may be present in. 
While this mis-match will seldom effect landscape ecologists, it may have implications for other data users.

@@

The formatting module has three functions.Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

@@

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.  
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.  
*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.

@@

The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online for preferred synonyms, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)` records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance where an outdated circumscription was applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.

In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.

Of `r n_spp` distinct species analysed BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling', na.rm = TRUE)`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling',  na.rm = TRUE)` unique species). 
In `r sum(epithet_diff$Situation=='Outdated', na.rm = TRUE)` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species). 
It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error. 
The final record was an egregious error where the order of the specific epithet and the genus name.

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.

Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users. 
While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
